---
title: "Vignette Title"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))

```{r, eval = FALSE}
set.seed(1)
iris_mat <- as.matrix(iris[1:4])
iris_km <- sapply(1:5, function(x) {
    km <- kmeans(iris_mat, centers = x, iter.max = 100, nstart = 10)
    km$cluster
})
colnames(iris_km) <- paste0("K", 1:5)
iris_clusts <- cbind(iris, iris_km)
```

```{r, eval = FALSE}
library("splatter")

set.seed(1)
sim <- splatSimulate(batchCells = 200, nGenes = 1000,
                     group.prob = c(0.4, 0.2, 0.2, 0.15, 0.05),
                     de.prob = c(0.1, 0.2, 0.05, 0.1, 0.05),
                     method = "groups")
sim_counts <- counts(sim)
```

```{r, eval = FALSE}
library("SC3")
library("scater")

sim_sc3 <- SingleCellExperiment(assays = list(counts = sim_counts))
rowData(sim_sc3)$feature_symbol <- rownames(sim_counts)
sim_sc3 <- normalise(sim_sc3)
sim_sc3 <- sc3(sim_sc3, ks = 1:8, biology = FALSE, n_cores = 1)
```

```{r, eval = FALSE}
library("Seurat")

sim_seurat <- new("seurat", raw.data = sim_counts,
                  cell.names = colnames(sim_counts),
                  meta.data = data.frame(cell = colnames(sim_counts)))
sim_seurat <- NormalizeData(sim_seurat, display.progress = FALSE)
sim_seurat <- FindVariableGenes(sim_seurat, do.plot = FALSE,
                                display.progress = FALSE)
sim_seurat <- ScaleData(sim_seurat, display.progress = FALSE)
sim_seurat <- RunPCA(sim_seurat, do.print = FALSE)
sim_seurat <- FindClusters(sim_seurat, dims.use = 1:6, resolution = 0,
                           save.SNN = TRUE, print.output = TRUE)
resolutions <- seq(0.1, 1, 0.1)
for (res in resolutions) {
   sim_seurat <- FindClusters(sim_seurat, resolution = res,
                              print.output = FALSE)
}
```

